---
title: "Practical Machine Learning Project"
author: "Shanthi Viswanathan"
date: "September 8, 2014"
output: html_document
---
### Synopsis
Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: 

 * exactly according to the specification (Class A), 
 * throwing the elbows to the front (Class B), 
 * lifting the dumbbell only halfway (Class C), 
 * lowering the dumbbell only halfway (Class D) and 
 * throwing the hips to the front (Class E).

The goal of this project is to predict the manner in which the participants did the exercise. Machine Learning is the process of using algorithms to learn from data.

Read more:http://groupware.les.inf.puc-rio.br/har#dataset

### The Data
In the study referenced above, the data was obtained by attaching sensors (inertial measurement units) to both study participants, and weights, to measure the motion as exercises were performed. 

In the final data set, available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv), each record consists of measurements obtained per person/exercise form in short span of time (0.5 - 2.5 seconds).

### Load needed libraries

```{r LoadLibraries}
library(caret);library(lattice);library(ggplot2)
set.seed(34567)
```

### Data Processing
First, we import the data and take an initial look at it:
```{r ReadData}
pml_training <- read.csv("pml-training.csv", na.strings=c("NA","","#DIV/0!"), strip.white=TRUE)
dim(pml_training)
```
There are 19622 records with 160 variables. 

```{r PlotData}
barplot(table(pml_training$classe), main="Bar plot for Class Type",  names.arg=c("Class A", "Class B", "Class C", "Class D", "Class E"), col=c("darkblue","red","darkgreen","yellow","magenta"), ylab="Count")
```
The variable we will be predicting on is **classe**, and the data is split up between the five **classes** A, B, C, D and E. Let us inspect the data to see if we can identify columns that are not needed for our prediction model.
```{r FindNAColumns}
na_data = sapply(pml_training, function(x) {sum(is.na(x))})
table(na_data)
```
There are only 60 columns with no null values and the remaining 100 columns have almost all missing values. We can remove these columns from our training data and take a look at the remaining columns.
```{r DiscardNAColumns}
bad_columns = names(na_data[na_data>19000])
new_training = pml_training[, !names(pml_training) %in% bad_columns]
```
The first seven columns are not related to the movement data and hence can be removed. 
```{r RemoveNonMovementColumns}
new_training = new_training [,-c(1:7)]
```
Split the data into training and test.
```{r PartitionData}
set.seed(0)
inTrain = createDataPartition(y=new_training$classe, p=0.7, list=FALSE)
training = new_training [inTrain,]
validate = new_training [-inTrain,]
dim(training)
```
#### Fit Random forest classification model
```{r RF, cache=TRUE}
ctrl <- trainControl(allowParallel=TRUE, method="cv", number=4)
modelRF <- train(classe ~ ., data=training, method="rf", trControl=ctrl)
modelRF
```
The model's accuracy is 1 and hence the out-of-sample error should be very minimal. 

#### Estimating out-of-sample error
We use the above model and apply on the *validation* data set.
```{r ValidateModel}
predictorRF <- predict(modelRF, validate)
cm <- confusionMatrix(predictorRF, validate$classe)
cm
```
The accuracy on validation data set is `r cm$overall[1][[1]]` which is very good. We can be **95% confident that the out-of-sample error of our model is between `r (1 - cm$overall[3][[1]])*100`% and `r (1 - cm$overall[4][[1]])*100`%**.

####Talk about Importance
```{r ImportanceRF}
plot(varImp (modelRF , scale = FALSE), top = 20)
trellis.par.set(caretTheme())
plot (modelRF , type = c("g", "o"))
```

### Predicting values for testing data set
```{r TestRF}
pml_test <- read.csv("pml-testing.csv", na.strings=c("NA","","#DIV/0!"),  strip.white=T)
pml_test <- pml_test[, !names(pml_test) %in% bad_columns]
pml_test <- pml_test [,-c(1:7)]
predict(modelRF, newdata=pml_test)
```

### References
1. Velloso, E., et al. Qualitative Activity Recognition of Weight Lifting Exercises. Available at http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201
2. Dataset source: http://groupware.les.inf.puc-rio.br/har